Discrete Quantum Theories and Computing

### 2019-02-22 dissertation.tex
dissertation.lyx "4.6.2 Discrete Grover Search: Nondeterministic" Replaced $\Delta$ by $V$ to avoid conflicting of symbols...

To solve Grover's problem, we represent the search states as $n$-qubit
states as usual. However, instead of the Deutsch black box~$U_{f}$,
$f$ is represented as the $N\times N$ ``phase rotation'' matrix
\begin{equation}
R=\begin{pmatrix}-1 & 0 & 0 & \cdots & 0\\
0 & 1 & 0 & \cdots & 0\\
0 & 0 & 1 & \cdots & 0\\
\vdots & \vdots & \vdots & \ddots & \vdots\\
0 & 0 & 0 & \cdots & 1
\end{pmatrix}\,,
\end{equation}
where the ``marked'' element is in the first position. Beside $R$,
we also need the $N\times N$ ``diffusion'' matrix
\begin{equation}
\Delta=\begin{pmatrix}1-\frac{N}{2} & 1 & 1 & \cdots & 1\\
1 & 1-\frac{N}{2} & 1 & \cdots & 1\\
1 & 1 & 1-\frac{N}{2} & \cdots & 1\\
\vdots & \vdots & \vdots & \ddots & \vdots\\
1 & 1 & 1 & \cdots & 1-\frac{N}{2}
\end{pmatrix}\,,
\end{equation}
where we have eliminated, in matrix~$\Delta$, the scaling factor
$\frac{2}{N}$ to enforce the requirement that all matrix coefficients
in our framework are integer-valued. By applying the transformation
$\Delta R$ repeatedly 
\begin{equation}
j=\round\left(\frac{\pi}{4\arccos\sqrt{1-\frac{1}{N}}}-\frac{1}{2}\right)\approx\round\left(\frac{\pi}{4}\sqrt{N}\right)
\end{equation}
times, we can find the target element~$\overline{0}$. In our context,
we also need to choose a prime number that is large enough to ensure
that all the numbers that occur during the calculation and after measurement
are within the transitively-ordered subrange.

Let's walk through the state in each iteration to make sure the algorithm
works. Because the probability amplitudes of $\ket{\overline{x}}$
are all the same for $\overline{x}\neq\overline{0}$, we can let $a_{l}$
be the probability amplitude of $\ket{\overline{0}}$, with $b_{l}$
the probability amplitude of each of the other possibilities, which
are all the same, after the operators $\Delta R$ is applied $l$
times. Beginning at $l=0$ with the information-less state and the
normalization scaled to integer values as usual, the state after $l$-th
iteration can be written as %
\begin{comment}
\todo{Verify the computation??}
\end{comment}
\begin{equation}
\begin{pmatrix}a_{l}\\
b_{l}\\
\vdots\\
b_{l}
\end{pmatrix}=\left(\Delta R\right)^{l}\begin{pmatrix}1\\
1\\
\vdots\\
1
\end{pmatrix}=\Delta R\begin{pmatrix}a_{l-1}\\
b_{l-1}\\
\vdots\\
b_{l-1}
\end{pmatrix}\,.
\end{equation}
We can solve the above iteration by the following recurrence relation
for the successive coefficients:
\begin{subequations}
\begin{align}
a_{0} & =1\,, & a_{l+1} & =\left(\frac{N}{2}-1\right)a_{l}+\left(N-1\right)b_{l}\,,\\
b_{0} & =1\,, & b_{l+1} & =\left(-1\right)a_{l}+\left(\frac{N}{2}-1\right)b_{l}\,.
\end{align}
\end{subequations}
We also know $\left|a_{j}\right|>\left|b_{j}\right|$, so we can estimate
an upper bound for the maximum cardinal probability as $\max\fnorm{a_{j}}\leq2\left(\frac{N}{2}\right)^{2j+1}$.
By applying Eq.~\eqref{eq:region} with $D=N=2^{n}$, we can estimate~$k$
using $k\geq8\left(\frac{N}{2}\right)^{2j+2}+1$. We can then pick
a prime $k$, and choose the $\pi\left(k\right)$-th prime in the
sequence represented by Table~\ref{pseq} guaranteeing that every
number we need for the computation is within the transitively ordered
range $F^{D}\left(k\right)$.

@@

Continuing with the $3$-qubit Grover example, we show how the cardinal
probabilities evolve to single out the target state. First, assume
that $f\left(\overline{x}\right)=1$ if and only if $\ket{\overline{x}}=\ket{0}\!\ket{0}\!\ket{0}$.
The initial information-less $8$-dimensional state vector evolves
under the application of $\Delta R$ as follows: %
\begin{comment}
\todo{Change the whole paragraph to a table?}
\end{comment}
\begin{equation}
\begin{pmatrix}1\\
1\\
\vdots\\
1
\end{pmatrix}\rightarrow\begin{pmatrix}10\\
2\\
\vdots\\
2
\end{pmatrix}\rightarrow\begin{pmatrix}44\\
-4\\
\vdots\\
-4
\end{pmatrix}\,.
\end{equation}
These states have differing field norm, so we cannot compare their
cardinal probability directly. Since states differ by only scalar
multiplication representing the same state, if we multiply the first
and second states by $16$ and $4$, respectively, their field norms
become the same value of $2048$. The now-consistently-normalized
states become 
\begin{equation}
\begin{pmatrix}16\\
16\\
\vdots\\
16
\end{pmatrix}\rightarrow\begin{pmatrix}40\\
8\\
\vdots\\
8
\end{pmatrix}\rightarrow\begin{pmatrix}44\\
-4\\
\vdots\\
-4
\end{pmatrix}\,.
\end{equation}
Therefore, the cardinal probabilities of measuring $\ket{0}\!\ket{0}\!\ket{0}$
in each state are 
\begin{align}
\cardp{256}{2048} &  & \cardp{1600}{2048} &  & \cardp{1936}{2048} & ,
\end{align}
while the cardinal probabilities of measuring the other states become
\begin{align}
\cardp{256}{2048} &  & \cardp{64}{2048} &  & \cardp{16}{2048} & .
\end{align}
We may thus conclude that the cardinal probability of measuring the
satisfying assignment of $f$ increases as we apply the diffusion
$\Delta$ and phase rotation $R$ matrices repeatedly.

>>>>

To solve Grover's problem, we represent the search states as $n$-qubit
states as usual. However, instead of the Deutsch black box~$U_{f}$,
$f$ is represented as the $N\times N$ ``phase rotation'' matrix
\begin{equation}
R=\begin{pmatrix}-1 & 0 & 0 & \cdots & 0\\
0 & 1 & 0 & \cdots & 0\\
0 & 0 & 1 & \cdots & 0\\
\vdots & \vdots & \vdots & \ddots & \vdots\\
0 & 0 & 0 & \cdots & 1
\end{pmatrix}\,,
\end{equation}
where the ``marked'' element is in the first position. Beside $R$,
we also need the $N\times N$ ``diffusion'' matrix
\begin{equation}
V=\begin{pmatrix}1-\frac{N}{2} & 1 & 1 & \cdots & 1\\
1 & 1-\frac{N}{2} & 1 & \cdots & 1\\
1 & 1 & 1-\frac{N}{2} & \cdots & 1\\
\vdots & \vdots & \vdots & \ddots & \vdots\\
1 & 1 & 1 & \cdots & 1-\frac{N}{2}
\end{pmatrix}\,,
\end{equation}
where we have eliminated, in matrix~$V$, the scaling factor $\frac{2}{N}$
to enforce the requirement that all matrix coefficients in our framework
are integer-valued. By applying the transformation $VR$ repeatedly
\begin{equation}
j=\round\left(\frac{\pi}{4\arccos\sqrt{1-\frac{1}{N}}}-\frac{1}{2}\right)\approx\round\left(\frac{\pi}{4}\sqrt{N}\right)
\end{equation}
times, we can find the target element~$\overline{0}$. In our context,
we also need to choose a prime number that is large enough to ensure
that all the numbers that occur during the calculation and after measurement
are within the transitively-ordered subrange.

Let's walk through the state in each iteration to make sure the algorithm
works. Because the probability amplitudes of $\ket{\overline{x}}$
are all the same for $\overline{x}\neq\overline{0}$, we can let $a_{l}$
be the probability amplitude of $\ket{\overline{0}}$, with $b_{l}$
the probability amplitude of each of the other possibilities, which
are all the same, after the operators $VR$ is applied $l$ times.
Beginning at $l=0$ with the information-less state and the normalization
scaled to integer values as usual, the state after $l$-th iteration
can be written as %
\begin{comment}
\todo{Verify the computation??}
\end{comment}
\begin{equation}
\begin{pmatrix}a_{l}\\
b_{l}\\
\vdots\\
b_{l}
\end{pmatrix}=\left(VR\right)^{l}\begin{pmatrix}1\\
1\\
\vdots\\
1
\end{pmatrix}=VR\begin{pmatrix}a_{l-1}\\
b_{l-1}\\
\vdots\\
b_{l-1}
\end{pmatrix}\,.
\end{equation}
We can solve the above iteration by the following recurrence relation
for the successive coefficients:
\begin{subequations}
\begin{align}
a_{0} & =1\,, & a_{l+1} & =\left(\frac{N}{2}-1\right)a_{l}+\left(N-1\right)b_{l}\,,\\
b_{0} & =1\,, & b_{l+1} & =\left(-1\right)a_{l}+\left(\frac{N}{2}-1\right)b_{l}\,.
\end{align}
\end{subequations}
We also know $\left|a_{j}\right|>\left|b_{j}\right|$, so we can estimate
an upper bound for the maximum cardinal probability as $\max\fnorm{a_{j}}\leq2\left(\frac{N}{2}\right)^{2j+1}$.
By applying Eq.~\eqref{eq:region} with $D=N=2^{n}$, we can estimate~$k$
using $k\geq8\left(\frac{N}{2}\right)^{2j+2}+1$. We can then pick
a prime $k$, and choose the $\pi\left(k\right)$-th prime in the
sequence represented by Table~\ref{pseq} guaranteeing that every
number we need for the computation is within the transitively ordered
range $F^{D}\left(k\right)$.

@@

Continuing with the $3$-qubit Grover example, we show how the cardinal
probabilities evolve to single out the target state. First, assume
that $f\left(\overline{x}\right)=1$ if and only if $\ket{\overline{x}}=\ket{0}\!\ket{0}\!\ket{0}$.
The initial information-less $8$-dimensional state vector evolves
under the application of $VR$ as follows: %
\begin{comment}
\todo{Change the whole paragraph to a table?}
\end{comment}
\begin{equation}
\begin{pmatrix}1\\
1\\
\vdots\\
1
\end{pmatrix}\rightarrow\begin{pmatrix}10\\
2\\
\vdots\\
2
\end{pmatrix}\rightarrow\begin{pmatrix}44\\
-4\\
\vdots\\
-4
\end{pmatrix}\,.
\end{equation}
These states have differing field norm, so we cannot compare their
cardinal probability directly. Since states differ by only scalar
multiplication representing the same state, if we multiply the first
and second states by $16$ and $4$, respectively, their field norms
become the same value of $2048$. The now-consistently-normalized
states become 
\begin{equation}
\begin{pmatrix}16\\
16\\
\vdots\\
16
\end{pmatrix}\rightarrow\begin{pmatrix}40\\
8\\
\vdots\\
8
\end{pmatrix}\rightarrow\begin{pmatrix}44\\
-4\\
\vdots\\
-4
\end{pmatrix}\,.
\end{equation}
Therefore, the cardinal probabilities of measuring $\ket{0}\!\ket{0}\!\ket{0}$
in each state are 
\begin{align}
\cardp{256}{2048} &  & \cardp{1600}{2048} &  & \cardp{1936}{2048} & ,
\end{align}
while the cardinal probabilities of measuring the other states become
\begin{align}
\cardp{256}{2048} &  & \cardp{64}{2048} &  & \cardp{16}{2048} & .
\end{align}
We may thus conclude that the cardinal probability of measuring the
satisfying assignment of $f$ increases as we apply the diffusion~$V$
and phase rotation~$R$ matrices repeatedly.
